---
title: "*biodb*: Using NCI CACTUS connector"
author: "Pierrick Roger"
date: "`r doc_date()`"
package: "`r pkg_ver('biodb')`"
abstract: |
  How to use the NCI CACTUS connector and its methods.
vignette: |
  %\VignetteIndexEntry{Using NCI CACTUS connector}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    toc: yes
    toc_depth: 4
    toc_float:
      collapsed: false
  BiocStyle::pdf_document: default
---

# Introduction

First we instantiate the package:
```{r}
mybiodb <- biodb::Biodb$new()
```

Then we create a connector to NCI CACTUS:
```{r}
cactus <- mybiodb$getFactory()$createConn('nci.cactus')
```

# Chemical Identifier Resolver web service

Here is an example of calling the Chemical Identifier Resolver for converting a SMILES into an InChI:
```{r}
cactus$wsChemicalIdentifierResolver(structid='C=O', repr='InChI')
```

# Convert CAS IDs

There are currently two methods in NCI CACTUS for converting from CAS IDs to InChI or InChI keys:
```{r}
cactus$convCasToInchi('87605-72-9')
cactus$convCasToInchikey('87605-72-9')
```

The conversion is made thanks to the Chemical Identifier Resolver web service.

# Closing biodb instance

Do not forget to terminate your biodb instance once you are done with it:
```{r}
mybiodb$terminate()
```
