---
title: "Accessing ChEBI entries"
author: "Pierrick Roger"
output:
  html_document:
    theme: null
package: biodb
abstract: |
  How to access entries in ChEBI database.
vignette: |
  %\VignetteIndexEntry{Accessing ChEBI entries}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# How to access entries in ChEBI database

## Introduction

Through the example of the ChEBI database, we are going to see how to access a database to access its entries.

First we instantiate the package:
```{r}
mybiodb <- biodb::Biodb$new()
```

Then we create a connector to ChEBI:
```{r}
chebi <- mybiodb$getFactory()$createConn('chebi')
```

## Requesting entries

Using accession numbers, we request entries from ChEBI:
```{r}
entries <- chebi$getEntry(c('2528', '17053', '15440'))
```

## Get entry fields

Getting the values of entry fields are done using `getFieldValue()` method.
Here we retrieve the SMILES fields for all ChEBI entries obtained previously:
```{r}
vapply(entries, function(e) e$getFieldValue('smiles'), FUN.VALUE = '')
```

## Get data frame

We can convert an entry into a data frame:
```{r}
entries[[1]]$getFieldsAsDataFrame()
```

Building a data frame for a list of entries is also possible:
```{r}
mybiodb$entriesToDataframe(entries, fields = c('accession', 'formula', 'molecular.mass', 'inchikey', 'kegg.compound.id'))
```

## Search by name and mass

Searching by name and/or mass is done through the `searchCompound()`.

Searching by name:
```{r}
chebi$searchCompound(name = 'aspartic', max.results = 3)
```

Searching by mass:
```{r}
chebi$searchCompound(mass = 133, mass.field = 'molecular.mass', mass.tol = 0.3, max.results = 3)
```

Searching by name and mass:
```{r}
ids <- chebi$searchCompound(name = 'aspartic', mass = 133, mass.field = 'molecular.mass', mass.tol = 0.3, max.results = 3)
```
Display results in data frame:
```{r}
mybiodb$entriesToDataframe(chebi$getEntry(ids), fields = c('accession', 'molecular.mass', 'name'))
```

## Closing biodb instance

Do not forget to terminate your biodb instance once you are done with it:
```{r}
mybiodb$terminate()
```
