---
title: "Details on biodb"
author: "Pierrick Roger"
date: "`r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('biodb')`"
abstract: |
  Details on general *biodb* usage and principles: configuration, architecture, ...
vignette: |
  %\VignetteIndexEntry{Details on general *biodb* usage and principles}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    toc: yes
    toc_depth: 4
    toc_float:
      collapsed: false
  BiocStyle::pdf_document: default
bibliography: references.bib 
---

# Introduction

In this document are presented different aspects of *biodb* in more details:
how it compares with other packages, the object oriented programming model
adopted, its architecture and how to configure it.

# History

<!--Plutôt le besoin d'un point du contexte international : besoin, existant, limites, ce que l'on propose -->
*biodb* originates in the need of merging chemical and biological data
downloaded from different public databases (*ChEBI*, *UniProt*, *miRBase*,
*NCBI*).
With the growing desire to connect to other databases (KEGG,
PeakForest, Massbank, ChemSpider, HMDB, local in-house database), features were
extended from getting entries and merging them, to mass spectra annotation,
mass spectra matching, pathway search, writing to local in-house database file
and framework for easy extension development.

With the last feature, the development framework, *biodb* were split into
different packages, one main package (*biodb*) holds the core with connectors
to local databases (in-house CSV file and SQLite databases), and several
extension packages contain each one or more connectors to related web databases
(KEGG, HMDB, ChEBI, ...).

The *biodb* core package is designed to automatically recognize and load
installed extension packages. No extension package is necessary for the core
package to work. And more important, the development and publication of a new
extension package can be made independently from the core package team.

# Comparison to other packages

<!-- webchem package -->
The R package most similar to *biodb* is *webchem* [@szocs2020_webchem]. Both packages
permit access to chemical databases and like *biodb*, *webchem* ensures the
respect of request frequency limit enacted by the different web database sites.
However several differences between the two packages make *biodb* special.

<!-- biology & MS -->
To begin with, the domain is different. While *webchem* is oriented toward
toxic compounds, *biodb* is centered around **biological compounds** and **mass
spectra**. Which leads to features specific to *biodb* like the access to
**protein** databases, **pathway** databases, and the **annotation** of mass spectra.

<!-- dev framework -->
Then, *biodb* has a distinctive approach concerning contributions. The authors
of *webchem* welcome contributors in a classical manner through the access to
their code repository.
With *biodb* we have decided to turn the package into a **development
framework** in order to simplify the development of new database connectors,
and see contributions to *biodb* as extension packages (i.e. plug-ins) instead
of code to be integrated into the main package repository. The consequence is a
greater independence for contributors, who can create and maintain their
extension package without the need to intervene into main package's code.

<!-- other features not in webchem -->
Finally *biodb* includes other specific features like a cache system that
avoids sending the same requests twice, the handling of in-house local
databases (CSV files and SQLite files), the search of compounds by mass and
name and a unified API that allows to use of generic methods (i.e. common to
between connectors) for entries access, entries search, spectra annotation and
spectra matching. See table \@ref(tab:features) for a list of main features.

```{r, child=c('includes/features_table.Rmd')}
```

<!-- TODO The need 
 - packages to connect to database are very different one from another (API and quality) ==> ...
-->
<!-- Facts -->
In the chemical and biological fields, the numerous databases available online are used on a daily bases by researchers across the world.
If manual online access through the web pages is usually well done on all public databases, this is not the case with automatic access from a program.
Solutions vary from the download of the whole database, mixed or homemade services, to well design web services using SOAP or REST.
We notice also a wide fluctuation in the formats used by databases for their entries and the results of requests: homemade formats, HTML, XML, JSON, CSV, SDF, etc.

<!-- Needs -->
For these reasons, developing and maintaining a code to access a database is not accessible to everyone and, for those who can, is time consuming.
Some R packages provide access to public databases, however, as of today, only one (<!-- TODO cite packages -->) gives access to several databases in the domain of toxic chemical compounds, and others (<!-- TODO cite packages -->) gives access to just one database server.
<!-- TODO so many databases, too few packages ==> need write homemade code to access db -->
<!-- TODO packages to do not offer access to all web services of a databases and to all data (fields) of entries -->
As a consequence many developers write their own code to access databases <!-- TODO to access the databases they want, and they do not share the code since packaging it into a library is time consuming -->
<!-- TODO what would be an improvement would be to facilitate the writing of code to access database and facilitate the sharing of this code as an R package -->

<!-- Comparison -->
<!-- TODO Reformulate. Move to vignette "details". -->
In comparison, *webchem* [@szocs2020_webchem], another R package providing access to chemical
databases, centers on toxic compounds, does not provide access to biological
and mass spectra databases, does not offer an open development framework for
facilitating extensions, does not use a cache system for web requests, and does
<!-- TODO forme positive: ce que biodb offre en plus -->
not include in-house databases.


<!-- Other packages -->
Other R packages exist to access 

<!-- TODO other packages -->
<!-- TODO

Linked R projects:

Python:
 * Python (Van Rossum et al. 2011) ecosystem 
   Van Rossum G, et al. (2011). Python Programming Language. URL http://www.python.org.

 * PubChempy (Swain 2017)
   Swain M (2017). “PubChemPy.” URL https://github.com/mcs07/PubChemPy.

 * ChemSpiPy (Swain 2018)
   Swain M (2018). “ChemSpiPy.” URL https://github.com/mcs07/ChemSpiPy.

 * CIRpy (Swain 2016)
   Swain M (2016). “CIRpy.” URL https://github.com/mcs07/CIRpy.

Web:
 * Chemical Translation Service (Wohlgemuth et al. 2010)
   Wohlgemuth G, Haldiya PK, Willighagen E, Kind T, Fiehn O (2010). “The Chemical Trans-
lation Service – A Web-Based Tool to Improve Standardization of Metabolomic Reports.”
Bioinformatics, 26(20), 2647–2648. doi:10.1093/bioinformatics/btq476.

Several KEGG packages. Some of them:
http://www.bioconductor.org/packages/release/bioc/html/KEGGREST.html
On graphs:
https://www.bioconductor.org/packages/release/data/annotation/html/KEGG.db.html
https://www.rdocumentation.org/packages/KEGGprofile/versions/1.14.0
https://www.rdocumentation.org/packages/KEGGgraph/versions/1.30.0

HMDB:
https://bioconductor.org/packages/release/bioc/html/hmdbQuery.html

UniProt:
https://cran.r-project.org/web/packages/UniprotR/

Package to export into MassBank:
http://www.bioconductor.org/packages/release/bioc/html/RMassBank.html

MetabolomeXchange is a web portal proposing web services that focus on metabolomics dataset search and retrieval, but is restricted to 3 databases and offers a limited set of services.
http://www.metabolomexchange.org/site/

BioServices is a Python module proposing a unified connection mechanism toward 25 biology web services.
https://academic.oup.com/bioinformatics/article/29/24/3241/194040/BioServices-a-common-Python-package-to-access

MAPI is a Java general framework for developing client applications for accessing web resources.
https://www.ncbi.nlm.nih.gov/pubmed/23311574

Having dependencies on other languages (=! C/C++) Not good for distribution 
-->

# Object oriented programming model

As you will see shortly in the next chapter, we use a call to a constructor to obtain an instance of the class `Biodb`. This means that *biodb* uses an object oriented programming (or OOP) model.

In R, you certainly already know the two classical OOP models *S3* and *S4*. *S4* is certainly the most used model, and is based on an original system of generic methods that can be specialized for each class.

Two more recent OOP models exist in R: *Reference Classes* (aka *RC* or *R5*) and *R6*. These two models are very similar. At the time *biodb* was created, only *R5* existed, and thus it has been chosen for the implementation of the package.

<!-- TODO Explain why R5 and not S4. -->

# Architecture

*biodb* uses an initialization/termination scheme. You must first initialize library by creating an instance:
```{r}
mybiodb <- biodb::Biodb()
```
And when you are done with the library, you have to terminate the instance explicitly:
```{r}
mybiodb$terminate()
```

When you are done working with your *biodb* instance, you must release it by
calling the `terminate()` method on it.

## Connector and entry classes

<!-- TODO
Connectors:
 * Share the same methods through the common super classes `BiodbConn`, `BiodbRemotedbConn`, ...
Entries:
 * Share the same methods through the common super class `BiodbEntry`.
 * the definition of fields are in common between database, thus the same field will have the same name in two entries of two different databases.
 Factory :
  * creates a new connector (createConn)
  * stores connectors in memory.
  * gives you back an existing connector (getConn)
  * creates entries
  * stores entries in memory (volatile cache).
  * gives you back existing entries (instead of reloading them)
  * you can delete entries from memory
  * you can delete connectors from memory
URL request scheduler:
 * run requests on a database server.
 * controls the frequency of requests for a database server.
 * Use the cache system to store results of requests.
 * Use cached request result instead of sending again the request.
Cache system :
a
 * stores entries on disk (persistence cache).
 * gives back entries instead of reloading them from database.
 * allows you to delete cached entries from disk.
 * allows you to delete all cached files for a database.
-->

## Management classes

<!-- TODO ==> put that in separate vignette about architecture. -->

Several class instances are attached to the *biodb* instance for managing different aspects of *biodb*: creating connectors, configuring *biodb*, accessing the cache system, etc.

### Factory

### Configuration

### Cache system

### Databases information

### Entry fields information

# Configuration

This configuration setting can be modified either directly inside R with the `BiodbConfig` class or beforehand by setting environment variables.

## BiodbConfig instance

The main way to tune *biodb* to your needs is to use the `BiodbConfig` single instance, accessible through the `biodb` instance:
```{r}
config <- mybiodb$getConfig()
```

If you look at the config instance, you will get a list of keys with their current values:
```{r}
config
```

## Keys information and value

Get all available configuration keys:
```{r}
config$getKeys()
```

Get description of a field:
```{r}
config$getDescription('cache.directory')
```

Get a field value:
```{r}
config$get('cache.directory')
```

If the field is boolean, you can use the following method instead:
```{r}
if (config$isEnabled('offline')) 'Biodb is running offline.' else 'Biodb is running online.'
```

To get a complete of all configuration keys and their description, call:
```{r}
config$listKeys()
```

## Setting a value

Set a field value:
```{r}
config$set('cache.directory', '~/my.biodb.cache')
config$get('cache.directory') # See modifications
```

If the field is boolean, you can use the following methods instead:
```{r}
config$enable('offline')    # set to TRUE
config$disable('offline')   # set to FALSE
```

## Default values

Get field default value:
```{r}
config$getDefaultValue('cache.directory')
```

## Environment variables

Environment variables can be used to overwrite default values.

To get the name of the environment variable associated with a particular key, call the following method:
```{r}
config$getAssocEnvVar('cache.directory')
```

# Closing biodb instance

Do not forget to terminate your biodb instance once you are done with it:
```{r}
mybiodb$terminate()
```

# References
