<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{Accessing database entries by accession number}
-->

# Accessing database entries by accession number

In this vignette, we will explain how to download and access entry information from databases.

First you need to create an instance of the `Biodb` class:
```{r}
# Create a Biodb instance
mybiodb <- biodb::Biodb$new()
```

Then you need to choose the database you want to access. For that you need to use a database ID in the form of a string. To get the list of available database IDs, just run:
```{r}
print(mybiodb$getDbsInfo()$getIds())
```

To obtain an entry from a database you have to call the `getEntry` method from the `BiodbFactory` class. The following example access the ChEBI database in order to retrieve the entry number 2528:
```{r}
# Request the entry 2528 in ChEBI
entry <- mybiodb$getFactory()$getEntry('chebi', id = '2528')
```

From an entry instance, you can access different values stored in fields. Here is an example to get back the entry number (or accession number):
```{r}
print(entry$getFieldValue('accession'))
```

We can have a list of all available fields for an entry:
```{r}
print(entry$getFieldNames())
```
