---
title: "Accessing database entries by accession number"
author: "Pierrick Roger"
output:
  html_document:
    theme: null
package: biodb
abstract: |
  How to retrieve entries from databases.
vignette: |
  %\VignetteIndexEntry{Accessing database entries by accession number}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

In this vignette, we will explain how to download and access entry information from databases.

First you need to create an instance of the `Biodb` class:
```{r}
mybiodb <- biodb::Biodb()
```

# Getting list of available databases

A database ID is required for choosing which database to access. To get the list of available database IDs, just run:
```{r}
mybiodb$getDbsInfo()$getIds()
```

# Getting an entry from the database
To obtain an entry from a database you have to call the `getEntry()` method from the `BiodbFactory` class. The following example access the KEGG Compound database in order to retrieve the entry number 2:
```{r}
entry <- mybiodb$getFactory()$getEntry('kegg.compound', id = 'C00005')
```

# Accessing entry fields
From an entry instance, you can access different values stored in fields. Here is an example to get back the entry number (or accession number) used to obtain this entry:
```{r}
entry$getFieldValue('accession')
```

We can have a list of all available fields for an entry:
```{r}
entry$getFieldNames()
```

You can also get a data frame of the field values with the method `getFieldsAsDataFrame()`:
```{r}
entry$getFieldsAsDataFrame()
```
