---
title: "*biodb*: In-house LCMS database"
author: "Pierrick Roger"
date: "`r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('biodb')`"
abstract: |
  How to define an in-house LCMS database and run annotation using it.
vignette: |
  %\VignetteIndexEntry{In-house LCMS database}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    toc: yes
    toc_depth: 4
    toc_float:
      collapsed: false
  BiocStyle::pdf_document: default
---

# Introduction

To load an in-house LCMS database, we are going to use the `mass.csv.file` database connector, which can be used for accessing an MS in-house database CSV file, an LCMS database CSV file, an MSMS database CSV file, or MSn, or a mix of all those different types.

We first create an instance of the `Biodb` class:
```{r}
mybiodb <- biodb::Biodb()
```

# Building an MS database from a data frame

We will use an extract of Massbank database for our example:
```{r}
fileUrl <- system.file("extdata", "massbank_extract_lcms_2.tsv", package='biodb')
```

We create the connector from the data frame, and will use this connector for all subsequent examples.
```{r}
conn <- mybiodb$getFactory()$createConn('mass.csv.file', url=fileUrl)
```

Two fields are missing inside this database: the MS level and unit used for chromatography retention times.
We define them using the `addField()` method:
```{r}
conn$addField('ms.level', 1)
conn$addField('chrom.rt.unit', 's')
```

# Searching for entries

Searching by M/Z:
```{r}
conn$searchMsEntries(mz.min = 145.96, mz.max = 147.13)
```

Searching with multiple M/Z:
```{r}
conn$searchMsEntries(mz = c(86.000002, 146.999997), mz.tol = 10, mz.tol.unit = 'ppm')
```

Searching by M/Z and RT:
```{r}
conn$searchMsEntries(mz.min = 72.78, mz.max = 73.12, rt = 45.5, rt.tol = 1, rt.unit = 's')
```

Other parameters can be used in `searchMsEntries()`:

 * MS level.
 * MS mode.
 * Minimum of relative intensity.
 * Matching of precursor peak.
 * Maximum of results.

# Peak annotation

Peak annotation can be done with `searchMsPeaks()` method.

We must first define the data frame of inputs:
```{r}
input <- data.frame(mz=c(73.01, 116.04, 174.2),
                    rt=c(79, 173, 79))
```

Then we run the annotation:
```{r}
conn$searchMsPeaks(input, mz.tol=0.1, rt.unit='s', rt.tol=10, match.rt=TRUE, prefix='match.')
```

# Closing biodb instance

Do not forget to terminate your biodb instance once you are done with it:
```{r}
mybiodb$terminate()
```
