---
title: "In-house mass spectra database"
author: "Pierrick Roger"
date: "`r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('biodb')`"
vignette: |
  %\VignetteIndexEntry{In-house LCMS database}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    toc: yes
    toc_depth: 4
    toc_float:
      collapsed: false
  BiocStyle::pdf_document: default
bibliography: references.bib
---

# Introduction

To load an in-house LCMS database, we are going to use the `mass.csv.file` database connector, which can be used for accessing an MS in-house database CSV file, an LCMS database CSV file, an MSMS database CSV file, or MSn, or a mix of all those different types.

We first create an instance of the `Biodb` class:
```{r}
mybiodb <- biodb::Biodb()
```

# Building an MS database from a data frame

We will use an extract of Massbank database for our example:
```{r}
fileUrl <- system.file("extdata", "massbank_extract_lcms_2.tsv", package='biodb')
```

We create the connector from the data frame, and will use this connector for all subsequent examples.
```{r}
conn <- mybiodb$getFactory()$createConn('mass.csv.file', url=fileUrl)
```

Two fields are missing inside this database: the MS level and unit used for chromatography retention times.
We define them using the `addField()` method:
```{r}
conn$addField('ms.level', 1)
conn$addField('chrom.rt.unit', 's')
```

# Searching for entries

Both `peak.mzexp` and `peak.mztheo` are defined inside the database.
The following call to get the field that will be used for matching will emit a warning because of this:
```{r}
conn$getMatchingMzField()
```

Setting the M/Z field to use:
```{r}
conn$setMatchingMzField('peak.mztheo')
```

Searching by M/Z:
```{r}
conn$searchMsEntries(mz.min=145.96, mz.max=147.13)
```

Searching with multiple M/Z:
```{r}
conn$searchMsEntries(mz=c(86.000002, 146.999997), mz.tol=10, mz.tol.unit='ppm')
```

Searching by M/Z and RT:
```{r}
conn$searchMsEntries(mz.min=72.78, mz.max=73.12, rt=45.5, rt.tol=1, rt.unit='s')
```

Other parameters can be used in `searchMsEntries()`:

 * MS level.
 * MS mode.
 * Minimum of relative intensity.
 * Matching of precursor peak.
 * Maximum of results.

# Peak annotation

Peak annotation can be done with `searchMsPeaks()` method.

We must first define the data frame of inputs:
```{r}
input <- data.frame(mz=c(73.01, 116.04, 174.2),
                    rt=c(79, 173, 79))
```

Then we run the annotation:
```{r}
conn$searchMsPeaks(input, mz.tol=0.1, rt.unit='s', rt.tol=10, match.rt=TRUE, prefix='match.')
```

# MS Annotation

## Annotation using an LCMS database

Here is the input data frame we want to annotate with both M/Z values and RT values in seconds:
```{r Set a list of M/Z values}
ms.tsv <- system.file("extdata", "ms.tsv", package='biodb')
mzdf <- read.table(ms.tsv, header=TRUE, sep="\t")
```

Annotation is also possible using an LCMS database.

For this example we use an in-house database saved as a TSV file:
```{r Instantiate an in-house LCMS database}
lcmsdb <- system.file("extdata", "massbank_extract_lcms_1.tsv", package="biodb")
massbank <- mybiodb$getFactory()$createConn('mass.csv.file', url=lcmsdb)
```
The database file was built using Massbank data. The accession numbers correspond to real Massbank entries.

Now that the database is loaded, we define some missing fields (MS level and RT unit):
```{r Set missing fields on database}
massbank$addField('ms.level', 1)
massbank$addField('chrom.rt.unit', 's')
```

We will now annotate our M/Z peaks, selecting a set of fields from the database to write in the output data frame:
```{r Annotate M/Z values with an in-house LCMS database}
massbank$searchMsPeaks(mzdf, mz.tol=1e-3, fields=c('accession', 'name', 'formula', 'chebi.id'), prefix='mydb.')
```

Annotation is also possible using retention times.
We then choose the chromatographic columns on which we want to match, and the set of fields we want in output:
```{r Set chromatographic columns and output fields}
chromColIds <- c('TOSOH TSKgel ODS-100V  5um Part no. 21456')
fields <- c('accession', 'name', 'formula', 'chebi.id', 'chrom.rt', 'chrom.col.id')
```

Finally we run the annotation on M/Z and RT values:
```{r Annotate using RT values}
massbank$searchMsPeaks(mzdf, mz.tol=1e-3, fields=fields, prefix='mydb.', chrom.col.ids=chromColIds, rt.unit='s', rt.tol=10, match.rt=TRUE)
```

For more details on the `searchMsPeaks()` method, please this the help page of `BiodbMassdbConn`.

# MS/MS

## Defining the database

We create the connector to a CSV file database built with MS2 spectra extracted from Massbank:
```{r}
db.tsv <- system.file("extdata", "massbank_extract_msms.tsv", package='biodb')
conn <- mybiodb$getFactory()$createConn('mass.csv.file', url=db.tsv)
```

## Searching for entries

We can search for entries containing certain M/Z values with the following method:
```{r}
conn$searchMsEntries(mz.min = 115, mz.max = 115.1, max.results = 5)
```

It is also possible to call it with a tolerance:
```{r}
conn$searchMsEntries(mz = 115, mz.tol = 0.1, mz.tol.unit = 'plain', max.results = 5)
```

## Spectrum matching

The spectrum to search for must be defined inside a data frame:
```{r}
spectrum <- data.frame(mz = c(286.1456, 287.1488, 288.1514), rel.int = c(100, 45, 18))
```

Then we call the `msmsSearch()` method:
```{r}
conn$msmsSearch(spectrum, precursor.mz = 286.1438, mz.tol = 0.1, mz.tol.unit = 'plain', ms.mode = 'pos')
```

# Closing biodb instance

Do not forget to terminate your biodb instance once you are done with it:
```{r}
mybiodb$terminate()
```
