---
title: "Merging entries coming from different databases"
author: "Pierrick Roger"
output:
  html_document:
    theme: null
package: biodb
abstract: |
  How to build a data frame containing entries coming from different databases.
vignette: |
  %\VignetteIndexEntry{Merging entries coming from different databases}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Merging entries coming from different databases

## Introduction

First we instantiate the package:
```{r}
mybiodb <- biodb::Biodb$new()
```

## Getting the entries

Request entries from ChEBI, using accession numbers:
```{r}
chebi.entries <- mybiodb$getFactory()$getEntry('chebi', id = c('2528', '17053'))
```

Request entries from Uniprot:
```{r}
uniprot.entries <- mybiodb$getFactory()$getEntry('uniprot', id = c('P02461', 'P08123', 'Q31611'))
```

Request entries from LIPID MAPS Structure:
```{r}
lipidmaps.entries <- mybiodb$getFactory()$getEntry('lipidmaps.structure', id = c('LMFA08040013'))
```

## Get a data frame of all these entries

We call then the `entriesToDataframe()` on all those entries, selectin the fields we want in the output:
```{r}
mybiodb$entriesToDataframe(c(chebi.entries, uniprot.entries, lipidmaps.entries), fields = c('accession', 'name', 'mass'))
```

## Closing biodb instance

Do not forget to terminate your biodb instance once you are done with it:
```{r}
mybiodb$terminate()
```
