---
title: "*biodb* Merging entries coming from different databases"
author: "Pierrick Roger"
date: "`r doc_date()`"
package: "`r pkg_ver('biodb')`"
abstract: |
  How to build a data frame containing entries coming from different databases.
vignette: |
  %\VignetteIndexEntry{Merging entries coming from different databases}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    toc: yes
    toc_depth: 4
    toc_float:
      collapsed: false
  BiocStyle::pdf_document: default
---

# Introduction

First we instantiate the package:
```{r}
mybiodb <- biodb::Biodb$new()
```

# Getting the entries

Request entries from ChEBI, using accession numbers:
```{r}
chebi.entries <- mybiodb$getFactory()$getEntry('chebi', id = c('2528', '17053'))
```

Request entries from Uniprot:
```{r}
uniprot.entries <- mybiodb$getFactory()$getEntry('uniprot', id = c('P02461', 'P08123', 'Q31611'))
```

# Get a data frame of all these entries

We call then the `entriesToDataframe()` on all those entries, selection the fields we want in the output:
```{r}
mybiodb$entriesToDataframe(c(chebi.entries, uniprot.entries), fields = c('accession', 'molecular.mass', 'monoisotopic.mass', 'inchikey', 'name'))
```

# Closing biodb instance

Do not forget to terminate your biodb instance once you are done with it:
```{r}
mybiodb$terminate()
```
