<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{Search MS dataabases}
-->

# Search MS dataabases

Create an instance of the `Biodb` class:
```{r}
mybiodb <- biodb::Biodb$new()
```

For this vignette, we will use the Massbank database. There are two other mass databases in biodb: Peakforest and in-house database. However the in-house database requires a vignette for itself, while the access Peakforest is currently restricted to the partners of the [MetaboHUB](http://www.metabohub.fr/home.html) project.

Create a connector to the MS database:
```{r}
conn <- mybiodb$getFactory()$createConn('massbank.jp')
```

## Getting M/Z values from the database

This feature is mainly useful for biodb tests purposes.

You can request a list of M/Z values from MS databases. Depending on the database, the list of M/Z values will be more or less exhaustive.

Getting a list of M/Z values:
```{r}
conn$getMzValues(max.results = 10)
```

You can restrict to a certain MS mode, ask for the peak to be a precursor peak, or even ask for an MS level:
```{r}
conn$getMzValues(max.results = 10, ms.level = 2, precursor = TRUE, ms.mode = 'pos')
```

## Search for spectra containing a peak

Here is how to search for spectra that contain a certain M/Z value:
```{r}
conn$searchMzRange(mz.min = 54, mz.max = 54.1, max.results = 5)
```

Another version is available that uses a tolerance instead of a range:
```{r}
conn$searchMzTol(mz = 54, mz.tol = 0.1, mz.tol.unit = 'plain', max.results = 5)
```
You can also set `mz.tol.unit` to `'ppm'`.

Both methods accept the following options:

Option        | Default | Description
------------- | ------- | --------------------------------------------------
`ms.mode`     |   `NA`  | Set to `'pos'` or `'neg'` to get only spectra from a certain MS mode.
`precursor`   | `FALSE` | When set to `TRUE`, the searched peak must be a precursor peak.
`ms.level`    |    0    | Set to an integer greater than 0 to get only spectra from this MS level.
`min.rel.int` |   `NA`  | The minimum of relative intensity required for the peak, in percentage from `0.0` to `100.0`.

## Search for MSMS spectra by spectrum matching

You can search a match of your MSMS spectrum inside the MSMS spectra of the database:
```{r}
# Define spectrum to match:
spectrum <- data.frame(mz = c(156.0124, 252.0446), rel.int = c(999, 158))

# Search for match:
conn$msmsSearch(spectrum, precursor.mz = 252.0448, mz.tol = 0.1, mz.tol.unit = 'plain', ms.mode = 'neg')
```
A data frame, ordered from highest score to lowest, is returned. It contains the following columns:

 * `id`: Database spectrum identifiers.
 * `score`: The matching score.
 * N columns `peak.#`: Each column corresponds to a peak of the searched spectrum (from first to last peak). A `-1` means that the peak has not been matched. An integer N greater than `0` means that the peak has been matched with the Nth peak of the database spectrum.
