---
title: "Creating a new field for entries."
author: "Pierrick Roger"
date: "`r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('biodb')`"
vignette: |
  %\VignetteIndexEntry{Creating a new field for entries.}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
abstract: |
  This vignettte explains the process to create a new *biodb* generic **entry field**, and how to **parse** its value for a particular database connector.
output:
  BiocStyle::html_document:
    toc: yes
    toc_depth: 4
    toc_float:
      collapsed: false
  BiocStyle::pdf_document: default
bibliography: references.bib
---

```{r, echo=FALSE}
source(system.file('vignettes_inc.R', package='biodb'))
```

# Introduction

In this vignette we will focus on creating a new *biodb* field to be used
inside an existing connector.
*biodb* fields are defined for all database connectors.
They are definitions of what types of data may be set inside *biodb* entry
objects.
Since they are shared by all connectors, they need to be defined without any
reference to a particular database.
However many of them are linked to a particular science or technological domain
(genetics, metabolomics, mass spectrometry, ...).

An entry field is like a type definition.
The definition is done at the top-level of *biodb*, and thus it not related to
any particular connector.
The definition includes: a name, a description, a class (integer, double,
character, logical), a cardinality (single value or vector), a list of allowed
values, a class (to group similar fields like "mass"), etc.

For a particular connector, when an entry object is created in memory, a file
containing the values is obtained from the database and a parsing is run in
order to extract those values and affect them to associated *biodb* entry
fields inside the *biodb* entry object.
Thus the parsing of the value of a *biodb* entry field is different for each
connector, while the *biodb* entry field is used by several different
connectors.

No *biodb* connector use all available *biodb* entry fields.
However it can happen that a connector does not implement the parsing of some
available data inside a database.
The reason is that, in most cases, the amount of available data, and the
diversity of it, inside a single entry would require an excessive amount of
coding.
As a consequence, we often restrict our development onto a subset of the
available data, in which we are interested.

When one particular data from the database is not present inside the entries of
the corresponding *biodb* connector, this means that no parsing has been
written for it inside the connector.
Moreover it could also mean that no *biodb* entry field is defined to handled
this particular type of data.
Fortunately, *biodb* offers you a way to correct dynamically, inside your code,
this shortage, creating a new *biodb* entry field if necessary and creating the
corresponding parsing of the data for the connector.

Follow the subsequent explanations in order to learn how to define a **new
parsing** of a value for a connector and assign it to an existing entry field,
and how to define a **new entry field**.

First we instantiate the package:
```{r}
mybiodb <- biodb::BiodbMain$new()
```

## Defining a new parsing of a field

Before going with the creation of a new field, we will look at different ways
of parsing a value for an existing *biodb* field that is not handled by a
connector.
Two connector cases will be used as examples: the `ChebiExConn` connector
defined for the
```{r, echo=FALSE, results='asis'}
make_vignette_ref('new_connector')
```
vignette and the `CompCsvFileConn` connector from the *biodb* package.

### Defining a parsing expression for a remote database connector (ChebiExConn)

The `ChebiExConn` class implements an example connector to the *ChEBI*
[@hastings2012_chebi] remote database.
See vignette
```{r, echo=FALSE, results='asis'}
make_vignette_ref('new_connector')
```
for the creation of this connector.

We load dynamically the definition of this connector inside *biodb* as
explained in the
```{r, echo=FALSE, results='asis'}
make_vignette_ref('new_connector')
```
vignette:
```{r}
chebiexDefFile <- system.file("extdata", "chebi_ex.yml", package='biodb')
connClass <- system.file("extdata", "ChebiExConn.R", package='biodb')
entryClass <- system.file("extdata", "ChebiExEntry.R", package='biodb')
source(connClass)
source(entryClass)
mybiodb$loadDefinitions(chebiexDefFile)
```
For our demonstration we will suppose this connector has been created by
somebody else, and we have no access to the implementation code.

We create a connector to this database:
```{r}
conn <- mybiodb$getFactory()$createConn('chebi.ex')
```
And get one entry:
```{r}
entryIds <- c('17001', '40304', '64679')
entriesDf <- mybiodb$entriesToDataframe(conn$getEntry(entryIds))
```
That you can see in table \@ref(tab:entriesChebiExTable).

```{r entriesChebiExTable, echo=FALSE, results='asis'}
# Prevent RMarkdown from interpreting @ character as a reference:
entriesDf$smiles <- vapply(entriesDf$smiles, function(s) paste0('`', s, '`'), FUN.VALUE='')
knitr::kable(head(entriesDf), "pipe", caption="Some entries from ChebiEx database. Their charge value is not visible become it was not parsed from database file.")
```

You will notice that no electrical charge is mentioned for the molecules in the table,
while it is present inside ChEBI database.
Let us choose one of the entries:
```{r}
id <- entryIds[[1]]
id
```
And get the ChEBI web page of this entry:
```{r}
conn$getEntryPageUrl(id)
```
Go on this page (
```{r, echo=FALSE, results='asis'}
cat("<", conn$getEntryPageUrl(id), ">", sep='')
```
) to check that the electrical charge information is indeed given by ChEBI (`Net Charge    0`).
To integrate this data inside the *biodb* entry, we need to extract it from the file returned by ChEBI.
When asked for an entry on its web service interface, ChEBI returns an XML file that *biodb* stores in its cache.
By calling the following method on your connector, you can get the path to the *biodb* cache file:
```{r}
conn$getCacheFile(id)
```
If you take a look to this file with your favourite editor, you will see the following text:
```
<charge>0</charge>
```
This the XML tag that stores the value of the electrical charge.
To extract values from XML, *biodb* uses the **XPath** query language.
In **XPath** language, the expression `//chebi:charge` means to get the value
inside the `charge` tag wherever it is (`//`) inside the tree structure of the
XML.
See [XPath Tutorial](http://www.w3schools.com/xsl/xpath_intro.asp) for an
introduction to XPath.
We need to give this **XPath** expression to the *biodb* instance, and explain
to which entry field the extracted value must be affected.
This is done by defining a small **YAML** file:
```{r}
chargeParsingDefFile <- system.file("extdata", "chebi_ex_charge_parsing.yml", package='biodb')
```
Whose content is as follow:
```{r, eval=FALSE, highlight=FALSE, code=readLines(chargeParsingDefFile)}
```
In this file we define a new parsing expression inside the `parsing.expr` section for the `chebi.ex` database connector.
The definition of the parsing expression consists of two values: the targeted *biodb* entry field (`charge`) and the *XPath* expression (`//chebi:charge`).

Now we just have to load this new definition:
```{r}
mybiodb$loadDefinitions(chargeParsingDefFile)
```
Delete the existing connector:
```{r}
mybiodb$getFactory()$deleteConn(conn)
```
Recreate the connector and reload the same entries:
```{r}
conn <- mybiodb$getFactory()$createConn('chebi.ex')
entriesDf <- mybiodb$entriesToDataframe(conn$getEntry(entryIds))
```
You can see in \@ref(tab:entriesChebiExWithChargeTable) that the electrical charge is now indicated for each entry.

```{r entriesChebiExWithChargeTable, echo=FALSE, results='asis'}
# Prevent RMarkdown from interpreting @ character as a reference:
entriesDf$smiles <- vapply(entriesDf$smiles, function(s) paste0('`', s, '`'), FUN.VALUE='')
knitr::kable(head(entriesDf), "pipe", caption="Some entries from ChebiEx database. Their charge value is now visible, since the parsing expression has been added to the connector.")
```

### Defining a parsing expression for a local database connector (CompCsvFileConn)

The `CompCsvFileConn` class implements a connector to a local CSV file database
of chemical compounds, as explained inside vignette
```{r, echo=FALSE, results='asis'}
make_vignette_ref('in_house_compound_db')
```
.
For a database stored inside a CSV file, the data parsing is very simple.
It consists in associating each *biodb* entry field with a column name.
By default *biodb* will define associations for each entry field whose name
is used for a column.
The columns whose names are not the names of existing *biodb* entry fields are
not associated and thus you cannot access their values from *biodb*.

If you want to access those values, you have the define manually the
associations, using the `setField()` method.

For our example we use an extract from ChEBI database as the input CSV database file:
```{r}
fileUrl <- system.file("extdata", "chebi_extract_with_unknown_column.tsv", package='biodb')
```
See table \@ref(tab:compDbTable) for the content of this file. 

```{r compDbTable, echo=FALSE, results='asis'}
compDbDf <- read.table(fileUrl, sep="\t", header=TRUE, quote="")
# Prevent RMarkdown from interpreting @ character as a reference:
compDbDf$smiles <- vapply(compDbDf$smiles, function(s) paste0('`', s, '`'), FUN.VALUE='')
knitr::kable(head(compDbDf), "pipe", caption="First lines of the compound database file.")
```
<!-- TODO préciser elecCharge n'est pas reconnu mais les autres oui
comment avoir la liste des noms de colonnes non reconnues
comment avoir la liste des noms de colonnes reconnues
comment avoir la liste des champs biodb définis
-->
In this file, the column name `elecCharge` will not be associated to any
*biodb* entry field.
Indeed, the *biodb* entry field the electrical charge of a molecule is
`charge`, not `elecCharge`.
Let us verify that.

We first create the connector to this CSV file:
```{r}
conn <- mybiodb$getFactory()$createConn('comp.csv.file', url=fileUrl)
```
And get the content of some of the entries:
```{r}
entriesDf <- mybiodb$entriesToDataframe(conn$getEntry(conn$getEntryIds()))
```
See table \@ref(tab:entriesTable) for the content of this entry. 
As you can see, no `charge` field is listed.

```{r entriesTable, echo=FALSE, results='asis'}
# Prevent RMarkdown from interpreting @ character as a reference:
entriesDf$smiles <- vapply(entriesDf$smiles, function(s) paste0('`', s, '`'), FUN.VALUE='')
knitr::kable(head(entriesDf), "pipe", caption="Some entries from the compound database.")
```

Now we call the method to define the new association:
```{r}
conn$setField('charge', 'elecCharge')
```
The first parameter is the name of the *biodb* entry field, the second the name
of the column inside the CSV file

The new column will now be parsed when getting the entry.
But before we must remove all entries from memory:
```{r}
conn$deleteAllEntriesFromVolatileCache()
```
And then reload the same entries again:
```{r}
entries2Df <- mybiodb$entriesToDataframe(conn$getEntry(conn$getEntryIds()))
```
See table \@ref(tab:entries2Table) for the content of this entry. 
A new data frame column is present, named `charge`.

```{r entries2Table, echo=FALSE, results='asis'}
# Prevent RMarkdown from interpreting @ character as a reference:
entries2Df$smiles <- vapply(entries2Df$smiles, function(s) paste0('`', s, '`'), FUN.VALUE='')
knitr::kable(head(entries2Df), "pipe", caption="Some entries from the compound database. They now show the newly parsed \"charge\" field.")
```

## Creating a new field and parsing its value

Sometimes you just do not need to parse some value for setting an existing
*biodb* field, but you need to get a value that does not correspond to any
defined *biodb* field. In this case, you need to define a **new field** alongside
defining your parsing.

For this demonstration we will use again the `ChebiExConn` connector example from the
```{r, echo=FALSE, results='asis'}
make_vignette_ref('new_connector')
```
vignette.

In the ChEBI database, each entry (i.e.: molecule) gets a score (a number of
stars) reflecting its curation status.
This field is not present inside the current `ChebiExConn` connector example.
Let us see that by displaying the content of some entries:
```{r}
conn <- mybiodb$getFactory()$getConn('chebi.ex')
entryIds <- c('17001', '40304', '64679')
entriesDf <- mybiodb$entriesToDataframe(conn$getEntry(entryIds))
```
See table \@ref(tab:entriesChebiEx2Table).

```{r entriesChebiEx2Table, echo=FALSE, results='asis'}
# Prevent RMarkdown from interpreting @ character as a reference:
entriesDf$smiles <- vapply(entriesDf$smiles, function(s) paste0('`', s, '`'), FUN.VALUE='')
knitr::kable(head(entriesDf), "pipe", caption="Some entries from ChebiEx database. There is no field that indicates the number of stars of an entry.")
```

In the XML entry content returned by the ChEBI server, this field is stored
inside the `entityStar` element as shown here:
```
<entityStar>3</entityStar>
```
You can check that directly inside the XML content of one of the entries, as
explained earlier. <!-- TODO make reference to chapter 1.1 - Rewrite commands ? -->

To get this number of stars we define the new field and its parsing expression
inside the following YAML file:
```{r}
nStarsDefFile <- system.file("extdata", "chebi_ex_stars_field.yml", package='biodb')
```
Here is its content:
```{r, eval=FALSE, highlight=FALSE, code=readLines(nStarsDefFile)}
```
You already know how to define the parsing expression inside the YAML file
The value of the XPath expression is a bit longer than for the electrical charge,
but the principle is the same.

What is new, is the `fields` section, in which we define the new fields.
The name of the field (`n_stars`) is used as a key inside the section.
Then several keys are used to define the field, see table
\@ref(tab:entryDefKeysTable) for a description of those keys.

Key                | Description
------------------ | ------------------------
alias              | Other possible names of the field.
description        | A description of the field.
class              | The R class. One of `integer`, `character`, `double`, `logical`.
type               | A name of a group for related fields. Existing ones are `name` and `mass`, but you can create your owns.
card               | The cardinality. Either `one` (single value) or `many` (vector).
case.insensitive   | If `true` then the value is case insensitive.
forbids.duplicates | If `true` and the cardinality is `many`, no duplicate values will be accepted. <!-- TODO Explain that duplicates are ignored silently ==> on error messages. -->
lower.case         | If `true`, the value will be put in lower case.
allowed.values     | If this vector is not empty, then only the values listed in this vector will be allowed for this field.
: (\#tab:entryDefKeysTable) The different keys used to define a field.

We can now load the new definition:
```{r}
mybiodb$loadDefinitions(nStarsDefFile)
```
Delete the existing connector:
```{r}
mybiodb$getFactory()$deleteConn(conn)
```
Recreate the connector and reload the same entries:
```{r}
conn <- mybiodb$getFactory()$createConn('chebi.ex')
entriesDf <- mybiodb$entriesToDataframe(conn$getEntry(entryIds))
```
See table \@ref(tab:entriesChebiEx2WithNstarsTable).
Now a column named `n_stars` indicates the number of stars for each entry in the data frame.

```{r entriesChebiEx2WithNstarsTable, echo=FALSE, results='asis'}
# Prevent RMarkdown from interpreting @ character as a reference:
entriesDf$smiles <- vapply(entriesDf$smiles, function(s) paste0('`', s, '`'), FUN.VALUE='')
knitr::kable(head(entriesDf), "pipe", caption="Some entries from ChebiEx database. Now there is a n_stars no field that indicates the number of stars of an entry.")
```

# Closing biodb instance

Do not forget to terminate your biodb instance once you are done with it:
```{r}
mybiodb$terminate()
```

# References
