---
title: "Creating a new field for entries."
author: "Pierrick Roger"
output:
  html_document:
    theme: null
package: biodb
abstract: |
  How to create a new field and use it to parse a value inside an entry content.
vignette: |
  %\VignetteIndexEntry{Creating a new field for entries.}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Creating a new field for entries.

## Introduction

In this vignette we will create a new field, and parse information from an entry content downloaded from a database for setting the value into the new field.

First we instantiate the package:
```{r}
mybiodb <- biodb::Biodb$new()
```

## Get an entry

## Look at the fields already parsed

## Look at the entry content

## Search for the information we need inside the content

Get the file path to the cache file containing the entry content.
Open the cache file into your favorite editor.
Look for the information you want to extract.

## Create the new field

## Create the parsing expression

<!--
	1. We need to move the constant parsing expressions into the database information file dbinfo.json
	2. We need to allow the definition of additional parsing expressions.
-->

## Forget/remove the loaded entry and remove it from the cache

## Get again the entry and see the new field

## Closing biodb instance

Do not forget to terminate your biodb instance once you are done with it:
```{r}
mybiodb$terminate()
```
