---
title: "Using an access token"
author: "Pierrick Roger"
output:
  html_document:
    theme: null
package: biodb
abstract: |
  How to use a token for accessing a database.
vignette: |
  %\VignetteIndexEntry{Using an access token}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Using an access token

## Setting your token

As of today, two databases require an access token: ChemSpider and Peakforest.

There are four ways to set an access token in *biodb*:

 * Through environment variables.
 * Through the `BiodbConfig` class instance.
 * When creating explicitly a connector instance.
 * Directly on the database information class (`BiodbDbInfo`) instance.

The environment variables and *biodb* configuration keys you must use to set your access tokens are:

Database            | *biodb* configuration key | Environment variable
------------------- | ------------------------- | -------------------------------------
PeakForest Mass     | peakforest.mass.token     | `BIODB_PEAKFOREST_MASS_TOKEN`
PeakForest Compound | peakforest.compound.token | `BIODB_PEAKFOREST_COMPOUND_TOKEN`
ChemSpider          | chemspider.token          | `BIODB_CHEMSPIDER_TOKEN`

To set your ChemSpider access token using the environment variables under UNIX/Linux systems:
```{bash}
export BIODB_CHEMSPIDER_TOKEN=01234567-89ab-cdef-0123-456789abcdef
```

The environment variable is used by the `BiodbConfig` class instance if available. However you can also use the `BiodbConfig` class instance directly:
```{r}
mybiodb <- biodb::Biodb$new()
mybiodb$getConfig()$set('chemspider.token', '01234567-89ab-cdef-0123-456789abcdef')
```

To set the token directly on the DbInfo instance:
```{r}
mybiodb <- biodb::Biodb$new()
dbsinfo <- mybiodb$getDbsInfo()
dbsinfo$get('chemspider')$setToken('01234567-89ab-cdef-0123-456789abcdef')
```

To set the token when instantiating the connector with the factory:
```{r}
mybiodb <- biodb::Biodb$new()
factory <- mybiodb$getFactory()
conn <- factory$createConn('chemspider', token = '01234567-89ab-cdef-0123-456789abcdef')
```

## Accessing the database

You should then be able to access the database normally.

## Closing biodb instance

Do not forget to terminate your biodb instance once you are done with it:
```{r}
mybiodb$terminate()
```
