<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{Configuring biodb}
-->

# Configuring biodb

In this vignette, we will explain how to configure biodb behaviour.

First you need to create an instance of the `Biodb` class:
```{r}
mybiodb <- biodb::Biodb$new()
```

Then get the configuration class:
```{r}
config <- mybiodb$getConfig()
```

Get all settable fields:
```{r}
print(config$getKeys())
```

Get description of a field:
```{r}
print(config$getDescription('cache.directory'))
```

Get a field value:
```{r}
print(config$get('cache.directory'))
```

Set a field value:
```{r}
config$set('cache.directory', '~/my.biodb.cache')
```

Get same field value again:
```{r}
print(config$get('cache.directory'))
```

Get field default value:
```{r}
print(config$getDefaultValue('cache.directory'))
```

If the value of a field is boolean, you can use the boolean methods instead of `get` and `set`:
```{r}
print(config$get('offline'))
config$enable('offline')    # set to TRUE
config$disable('offline')   # set to FALSE
print(if (config$isEnabled('offline')) 'Mode offline is ON.' else 'Mode offline is OFF.')
```
