<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{Configuring biodb}
-->

# Configuring biodb

In this vignette, we will explain how to configure biodb behaviour.

First you need to create an instance of the `Biodb` class:
```{r}
mybiodb <- biodb::Biodb$new()
```

Then get the configuration class:
```{r}
config <- mybiodb$getConfig()
```

## Getting and setting a value

Get all settable fields:
```{r}
config$getKeys()
```

Get description of a field:
```{r}
config$getDescription('cache.directory')
```

Get a field value:
```{r}
config$get('cache.directory')
```

Set a field value:
```{r}
config$set('cache.directory', '~/my.biodb.cache')
```

Get same field value again:
```{r}
config$get('cache.directory')
```

Get field default value:
```{r}
config$getDefaultValue('cache.directory')
```

If the value of a field is boolean, you can use the boolean methods instead of `get` and `set`:
```{r}
config$get('offline')
config$enable('offline')    # set to TRUE
config$disable('offline')   # set to FALSE
if (config$isEnabled('offline')) 'Biodb is running offline.' else 'Biodb is running online.'
```
