<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{Initializing biodb}
-->

# Initializing biodb

First of all, before using *biodb*, you need to create an instance of the main class `Biodb`. This is done by calling the constructor of the class through the `new()` method. The OOP (Object Oriented Programming) system used in the *biodb* package is the **R5** system also called **Reference Classes**. In this system, the created objects are not copied, but their reference are copied. This means than when you pass an instance to a function, that function is able to modify the instance.

To create an instance of the `Biodb` class, call the `new()` method:
```{r}
mybiodb <- biodb::Biodb$new()
```
This is will create a `Biodb` instance and put its reference into the `mybiodb` variable. Calling `new()` without any argument configures the instance with the default values.

## Observers

By default, the `Biodb` instance will be configured with a console logger (instance of the class `BiodbLogger`), which will print all *biodb* messages to the console.
You can disable this default logger by setting the `logger` argument to false:
```{r}
mybiodb <- biodb::Biodb$new(logger = FALSE)
```
In this case, only the normal R warnings and errors will be printed.

### Setting your own instance of the `BiodbLogger` class

You can also set your own instance of `BiodbLogger`, using the `observers` argument, in order for instance to redirect the output to a file:
```{r}
mybiodb <- biodb::Biodb$new(logger = FALSE, observers = biodb::BiodbLogger$new(file = 'mybiodb.log'))
```

If you want both have the logger output written to the console and into a file, just define two instances of the `BiodbLogger` class:
```{r}
mybiodb <- biodb::Biodb$new(logger = FALSE, observers = list(biodb::BiodbLogger$new(), biodb::BiodbLogger$new(file = 'mybiodb2.log')))
```

See the documentation of the `BiodbLogger` class for more details.

### Defining your own observers

If you wish, you can also define your own observers that inherit from the `BiodbObserver` class, and set them through the `observers` argument.

### Setting observers after the instance has been created

It is also possible to add an observer during the life of your *biodb* instance:
```{r}
mybiodb$addObservers( list(biodb::BiodbLogger$new(), biodb::BiodbLogger$new(file = 'mybiodb3.log')))
```

### Getting a list of current observers

Here is how to obtain a list of all observers currently defined:
```{r}
mybiodb$getObservers()
```

## The singleton classes

Several singleton classes are defined in *biodb*. Their instances are all accessible from the `Biodb` instance you've created.

To get the factory instance:
```{r}
factory <- mybiodb$getFactory()
```

To get the configuration instance:
```{r}
config <- mybiodb$getConfig()
```

To get the cache system instance:
```{r}
cache <- mybiodb$getCache()
```

To get the database information instance:
```{r}
dbsinfo <- mybiodb$getDbsInfo()
```

To get the entry fields instance:
```{r}
entry.fields <- mybiodb$getEntryFields()
```
